Accesible, realistic genome simulation with selection using \stdpopsim

### 2025-03-10 analysis2_manuscript.tex
abstract

\author[2]{Maria Izabel A. Cavassim}
% \author[5,7]{Jeff Adrion}
% \author[5]{Saurabh Belsare}
% \author[8]{Arjun Biddanda}
% \author[5]{Victoria Caudill}
% \author[9]{Jean Cury}
% \author[10]{Ignacio Echevarria}
% \author[11]{Benjamin C. Haller}
% \author[12,13]{Ahmed R. Hasan}
\author[14,15]{Xin Huang}
% \author[16]{Leonardo Nicola Martin Iasi}
% \author[17]{Ekaterina Noskova}
% \author[18]{Jana Obšteter}
% \author[19]{Vitor Antonio Corrêa Pavinato}
% \author[20,21]{Alice Pearson}
% \author[22,23]{David Peede}
% \author[24]{Manolo F. Perez}
\author[5]{Chris C. R. Smith}
% \author[25]{Jeffrey P. Spence}
% \author[5]{Anastasia Teterina}

@@

\section*{Abstract}
    \label{abstract}
    Natural selection is a fundamental evolutionary force that shapes patterns of genetic variation across species. 
    However, simulating realistic models that incorporate both selection and complex demographic histories
    is challenging, limiting our ability to benchmark statistical methods aimed at detecting selection and explore theoretical predictions.
    \stdpopsim is a tool for facilitating the simulation of a contiually expanding catalog of population genetic models.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of natural selection, including background selection, selective sweeps, and arbitrary distributions of fitness effects (DFE).
    This extension maintains \stdpopsim's core principles of reproducibility and standardization while adding support
    for species-specific genomic annotations and published DFE estimates. 
    We demonstrate the utility of this framework by benchmarking methods for demographic inference,
    DFE estimation, and selective sweep detection across different species and scenarios. 
    Our results demonstrate the robustness of demographic inference methods to selection on linked sites, 
    reveal the sensitivity of DFE inference methods to model assumptions, 
    and show how genomic features like recombination rate and the density of functional sequence influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible framework.

@@

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community. %maybe add citations here?
    While that is so, the original version of \stdpopsim did not include
    models of selection, which is a major limitation for more empirically motivated
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation\citep[e.g.][]{thornton2019polygenic,hartfield2024polygenic}, the effects of selective sweeps \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, and the impact of
    background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Furthermore the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Ideally, one should be able
    to model all of these features simultaneously using estimates and annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background,rodrigues2024shared}.

@@

    % We used three software packages to infer the distribution of fitness effects (DFE) from genetic data:
    % \dadicli \citep{Huang2023,kim2017inference}, \polydfe \citep{tataru2020polydfe}, and \grapes \citep{galtier2016adaptive}.
    % Each software package was run using the same set of simulations.
    % For these analyses we used the exon annotations from the human and vaquita porpoise genomes respectively.
    % A key distinction between our analyses and those done in practice
    % is that we have complete knowledge of which sites are under selection in the simulation,
    % thus our inferences is a best case scenario. Empirical inference of the DFE is a much
    % harder problem as some mutations assumed to be free of selection
    % may in reality be under selection. As above, all methods were run using default settings.
    \igcomment{I edited the description below written by Silas. Please check to make sure it's still accurate.}
    As described above, each DFE-inference method scales selection coefficients slightly differently.
    We thus rescaled each program's output to obtain a selection coefficient ($s$) consistent with the one used by SLiM.
    The output of \grapes  was scaled by a factor of  $\mu / \theta$,
    and the output of \polydfe was saled by $2 \mu / \theta$, 
    where $\mu$ is the assumed mutation rate and 
    $\theta$ is the population-scaled mutation rate estimated by each method ($\theta = 4N_{e}\mu$).
    For \dadicli, the gamma distribution is parameterized using the shape and rate ($\alpha$ and $\nu$, respectively). 
    The mean value of the distribution, which is $s$ is transformed from $\alpha$ and $\nu$ and rescaled as $\alpha * \nu / (\theta / (4 \mu L p))$,
    where $L$ is the total number of sites available for the analysis,
    and $p$ is the proportion of sites under selection.

>>>>

\author[2]{Maria Izabel A. Cavassim}
% \author[5,7]{Jeff Adrion}
% \author[5]{Saurabh Belsare}
% \author[8]{Arjun Biddanda}
% \author[5]{Victoria Caudill}
% \author[9]{Jean Cury}
% \author[10]{Ignacio Echevarria}
% \author[11]{Benjamin C. Haller}
% \author[12,13]{Ahmed R. Hasan}
\author[14,15]{Xin Huang}
\author[5]{Kevin Korfmann}
% \author[16]{Leonardo Nicola Martin Iasi}
% \author[17]{Ekaterina Noskova}
% \author[18]{Jana Obšteter}
% \author[19]{Vitor Antonio Corrêa Pavinato}
% \author[20,21]{Alice Pearson}
% \author[22,23]{David Peede}
% \author[24]{Manolo F. Perez}
\author[5]{Chris C. R. Smith}
% \author[25]{Jeffrey P. Spence}
% \author[5]{Anastasia Teterina}

@@

\section*{Abstract}
    \label{abstract}
    Natural selection is a fundamental evolutionary force that shapes patterns of genetic variation across species. 
    However, simulating realistic models that incorporate both selection and complex demographic histories
    is challenging, limiting our ability to benchmark statistical methods aimed at detecting selection and to explore theoretical predictions.
    \stdpopsim, a simulation library that provides an extensive catalog of population genetic models,
    aims to provide a standardized interface for simulating realistic models that incorporate both selection and complex demographic histories.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of natural selection, including background selection, selective sweeps, and arbitrary distributions of fitness effects (DFE).
    This extension maintains \stdpopsim's core principles of reproducibility and standardization while adding support
    for species-specific genomic annotations and published DFE estimates. 
    We demonstrate the utility of this framework by benchmarking methods for demographic inference,
    DFE estimation, and selective sweep detection across different species and scenarios. 
    Our results demonstrate the robustness of demographic inference methods to selection on linked sites, 
    reveal the sensitivity of DFE inference methods to model assumptions, 
    and show how genomic features like recombination rate and the density of functional sequence influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful, new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible framework.

@@

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community. %maybe add citations here?
    While that is so, the original version of \stdpopsim did not include
    models of selection, which is a major limitation for more empirically motivated
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation\citep[e.g.][]{thornton2019polygenic,hartfield2024polygenic}, the effects of selective sweeps \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, and the impact of
    background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Furthermore the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Moreover, one should be able
    to model all of these features simultaneously using estimates and annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background,rodrigues2024shared}.

@@

    % We used three software packages to infer the distribution of fitness effects (DFE) from genetic data:
    % \dadicli \citep{Huang2023,kim2017inference}, \polydfe \citep{tataru2020polydfe}, and \grapes \citep{galtier2016adaptive}.
    % Each software package was run using the same set of simulations.
    % For these analyses we used the exon annotations from the human and vaquita porpoise genomes respectively.
    % A key distinction between our analyses and those done in practice
    % is that we have complete knowledge of which sites are under selection in the simulation,
    % thus our inferences is a best case scenario. Empirical inference of the DFE is a much
    % harder problem as some mutations assumed to be free of selection
    % may in reality be under selection. As above, all methods were run using default settings.
    \igcomment{I edited the description below written by Silas. Please check to make sure it's still accurate.}
    As described above, each DFE-inference method scales selection coefficients slightly differently.
    We thus rescaled each program's output to obtain a selection coefficient ($s$) consistent with the one used by SLiM.
    The output of \grapes  was scaled by a factor of  $\mu / \theta$,
    and the output of \polydfe was saled by $2 \mu / \theta$, 
    where $\mu$ is the assumed mutation rate and 
    $\theta$ is the population-scaled mutation rate estimated by each method ($\theta = 4N_{e}\mu$).
    \igcomment{Do we really multiple $\alpha$ by $\nu$ in rescaling for \dadicli?}
    For \dadicli, we report $\alpha * \nu / (\theta / (4 \mu L p))$, where
    $\alpha$ and $\nu$ are the inferred shape and rate parameters of the \dadicli's inferred gamma distribution,
    $L$ is the total number of sites available for the analysis,
    $p$ is the proportion of sites under selection.
